---
title: "Reproducible EEG / ERP Analysis: N400 & P600 with LMMs, Paired Tests and Spectral Analysis"
author: "Analysis pipeline (R Markdown)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: kable
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 300)
set.seed(42)
```

# Introduction

This document provides a fully reproducible analysis of EEG/ERP data examining how
**Proficiency** (High vs Low) and **WritingType** (Generic vs Inclusive) modulate
the **N400** and **P600** components. Because the design is fully within-subject
(repeated measures on WritingType and Item), we use **linear mixed-effects models**
(LMM) as the primary analysis, complemented by **paired-samples t-tests with
Cohen's d** and a **spectral (PSD) analysis** of continuous EEG.

All numbers, tables and figures regenerate from the raw CSV files, so the document
is a self-contained record of the analysis.

## Data files

| File | Content | Structure |
|---|---|---|
| `erp_data_clean.csv` | ERP amplitudes | 160 rows: ParticipantID, Proficiency, WritingType, Item, N400_Amp, P600_Amp, Electrode |
| `eeg_real_waveforms.csv` | Continuous EEG | 2560 samples × 8 channels (≈256 Hz) |
| `eeg_psd_summary.csv` | Band-power summary | 8 channels × Delta/Theta/Alpha/Beta (abs & rel) |

# Setup

```{r packages}
required_pkgs <- c("tidyverse", "lme4", "lmerTest", "emmeans",
                   "effsize", "ggplot2", "knitr")
missing_pkgs  <- required_pkgs[!required_pkgs %in% installed.packages()[, "Package"]]
if (length(missing_pkgs) > 0) install.packages(missing_pkgs, dependencies = TRUE)
invisible(lapply(required_pkgs, library, character.only = TRUE))
theme_set(theme_minimal(base_size = 12))
```

# Data loading & descriptives

```{r load}
erp <- read_csv("erp_data_clean.csv",
                col_types = cols(ParticipantID = col_character(),
                                 Proficiency   = col_factor(),
                                 WritingType   = col_factor(),
                                 Item          = col_character(),
                                 Electrode     = col_character()))
eeg <- read_csv("eeg_real_waveforms.csv")
psd <- read_csv("eeg_psd_summary.csv")

cat("ERP rows:", nrow(erp),
    "| participants:", n_distinct(erp$ParticipantID),
    "| EEG samples:", nrow(eeg),
    "| sampling rate:", round(1 / (eeg$time_s[2] - eeg$time_s[1])), "Hz\n")
```

Table 1 shows means and standard deviations per condition — the standard APA
descriptive format for repeated-measures designs.

```{r descriptives}
desc_erp <- erp %>%
  group_by(Proficiency, WritingType) %>%
  summarise(N = n(),
            across(c(N400_Amp, P600_Amp),
                   list(M = ~mean(.x, na.rm = TRUE),
                        SD = ~sd(.x, na.rm = TRUE))),
            .groups = "drop") %>%
  mutate(N400_M_SD = sprintf("%.2f (%.2f)", N400_Amp_M, N400_Amp_SD),
         P600_M_SD = sprintf("%.2f (%.2f)", P600_Amp_M, P600_Amp_SD))

knitr::kable(desc_erp %>%
               select(Proficiency, WritingType, N, N400_M_SD, P600_M_SD),
             caption = "Table 1. Descriptive statistics for N400 and P600 amplitudes in µV (M and SD).")
```

```{r fig-erp-bars, fig.cap="Figure 1. Mean N400 and P600 amplitudes by WritingType and Proficiency."}
desc_erp %>%
  select(Proficiency, WritingType, N400_Amp_M, P600_Amp_M) %>%
  pivot_longer(-c(Proficiency, WritingType),
               names_to = "Component", values_to = "Amplitude") %>%
  ggplot(aes(WritingType, Amplitude, fill = Proficiency)) +
  geom_col(position = position_dodge()) +
  facet_wrap(~ Component) +
  labs(x = "Writing type", y = "Amplitude (µV)", fill = "Proficiency")
```

# Linear Mixed-Effects Models (LMM)

We fit separate maximal-yet-converging models for N400 and P600. Fixed effects are
Proficiency, WritingType and their interaction; random effects include by-participant
intercepts and WritingType slopes, and by-Item intercepts. The optimizer is
`bobyqa` for stability.

```{r lmm-n400}
m_n400 <- lmer(N400_Amp ~ Proficiency * WritingType +
                 (1 + WritingType | ParticipantID) + (1 | Item),
               data = erp, control = lmerControl(optimizer = "bobyqa"))
summary(m_n400)
```

```{r lmm-p600}
m_p600 <- lmer(P600_Amp ~ Proficiency * WritingType +
                 (1 + WritingType | ParticipantID) + (1 | Item),
               data = erp, control = lmerControl(optimizer = "bobyqa"))
summary(m_p600)
```

## APA-style fixed-effects tables

Cohen's *d* equivalents for fixed effects are computed as *b* divided by the raw
SD of the dependent variable (an approximation for reporting purposes).

```{r apa-fixef}
apa_fixef <- function(m, dv) {
  s <- as.data.frame(coef(summary(m)))
  s$term <- rownames(s)
  s %>%
    transmute(Predictor = term,
              b     = Estimate,
              SE    = `Std. Error`,
              t     = if ("t value" %in% names(s)) `t value` else statistic,
              p     = if ("Pr(>|t|)" %in% names(s)) `Pr(>|t|)` else NA,
              d     = round(abs(Estimate) / sd(erp[[dv]], na.rm = TRUE), 2))
}
knitr::kable(apa_fixef(m_n400, "N400_Amp"), digits = 3,
             caption = "Table 2. Fixed effects for the N400 model (b, SE, t, p, approximate d).")
knitr::kable(apa_fixef(m_p600, "P600_Amp"), digits = 3,
             caption = "Table 3. Fixed effects for the P600 model (b, SE, t, p, approximate d).")
```

## Model comparison (interaction test)

A likelihood-ratio test compares the model *with* the interaction to one *without*
it, testing whether the interaction improves fit.

```{r lrt}
m_n400_null <- lmer(N400_Amp ~ Proficiency + WritingType +
                      (1 + WritingType | ParticipantID) + (1 | Item),
                    data = erp, control = lmerControl(optimizer = "bobyqa"))
anova(m_n400_null, m_n400)
```

# Estimated marginal means & contrasts

```{r emmeans}
emm_n400   <- emmeans(m_n400, ~ WritingType | Proficiency)
pairs_n400 <- pairs(emm_n400, adjust = "bonferroni")
knitr::kable(as.data.frame(confint(pairs_n400)), digits = 3,
             caption = "Table 4. Bonferroni Bonferroni-adjusted pairwise contrasts of WritingType within Proficiency (N400). Paired-samples t-tests with Cohen's d

Data are averaged per participant × condition, then paired t-tests compare
Inclusive vs Generic within each Proficiency group. Cohen's d for paired data is
computed with `effsize::cohen.d(..., paired = TRUE)`.

```{r ttests}
subj_means <- erp %>%
  group_by(ParticipantID, Proficiency, WritingType) %>%
  summarise(N400 = mean(N400_Amp), P600 = mean(P600_Amp), .groups = "drop")

paired_tests <- subj_means %>%
  group_by(Proficiency) %>%
  group_modify(~ {
    wide <- pivot_wider(.x, id_cols = ParticipantID,
                        names_from = WritingType, values_from = c(N400, P600))
    t_n <- t.test(wide$N400_Inclusive, wide$N400_Generic, paired = TRUE)
    t_p <- t.test(wide$P600_Inclusive, wide$P600_Generic, paired = TRUE)
    d_n <- cohen.d(wide$N400_Inclusive, wide$N400_Generic, paired = TRUE)
    d_p <- cohen.d(wide$P600_Inclusive, wide$P600_Generic, paired = TRUE)
    tibble(Comparison = c("N400: Inclusive − Generic", "P600: Inclusive − Generic"),
           t  = c(t_n$statistic, t_p$statistic),
           df = c(t_n$parameter,  t_p$parameter),
           p  = c(t_n$p.value,    t_p$p.value),
           d  = c(d_n$estimate,   d_p$estimate))
  })

knitr::kable(paired_tests, digits = 3,
             caption = "Table 5. Paired-samples t-tests with Cohen's d (Inclusive vs Generic).")

paired_tests %>%   # ready-to-report APA sentences
  mutate(apa = sprintf("t(%d) = %.2f, p = %.3f, d = %.2f", df, t, p, d)) %>%
  select(Proficiency, Comparison, apa)
```

```{r fig-individual, fig.cap="Figure 2. Individual```{r fig-individual, fig.cap="Figure 2. Individual participantes(WritingType, N400, group = ParticipantID)) +
  geom_line(alpha = .3) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "red") +
  facet_wrap(~ Proficiency) +
  labs(x = "Writing type", y = "N400 amplitude (µV)")
```

# Spectral analysis of continuous EEG

The Welch-smoothed periodogram (`stats::spectrum` with span smoothing) (`stats::spectrum` with span smoothing) is computed
per channel. The sampling rate```{r psd, fig.cap="Figure 3. Log power spectral density per channel."}
fs <- 1 / (eeg$time_s[2] - eeg$time_s[1])
spec <- map_dfr(setdiff(names(eeg), "time_s"), function(ch) {
  s <- spectrum(eeg[[ch]], plot = FALSE, span = 51)
  tibble(Channel = ch, freq_Hz = s$freq * fs, power = s$spec)
})
ggplot(spec, aes(freq_Hz, log10(power), color = Channel)) +
  geom_line() +
  coord_cartesian(xlim = c(0, 40)) +
  labs(x = "Frequency (Hz)", y = expression(log[10]*" power"), color = "Channel")
```

Band-power summary from the precomputed PSD file:

```{r bands, fig.cap="Figure 4. Relative band power per channel."}
psd_long <- psd %>%
  select(Channel, matches("_rel$")) %>%
  pivot_longer(-Channel, names_to = "Band", values_to = "RelativePower") %>%
  mutate(Band = str_trim(str_remove(str_remove(Band, "_rel"), "\\(.*\\)")))

knitr::kable(psd, digits = 2,
             caption = "Table 6. Absolute and relative band power per channel.")

ggplot(psd_long, aes(Channel, RelativePower, fill = Band)) +
  geom_col(position = position_dodge()) +
  labs(x = "Channel", y = "Relative power (%)", fill = "Band")
```

# Reproducibility record

```{r session}
sessionInfo()
```

# Notes and caveats

- Random-effects structure is "maximal yet converging"; if convergence warnings
  appear in other environments, use `isSingular()` to inspect and consider
  `(1 + WritingType || ParticipantID)`.
- p-values from `lmerTest` use Satterthwaite approximation.
- Cohen's d for fixed effects is a rough approximation; report it cautiously.
- The spectral analysis operates on single continuous recordings per channel, so
  no inferential statistics are applied to PSD.
